### 2024-10-21 starter_folder-main/paper/paper.qmd
Update paper.qmd
Just to have something saved for now

---
title: "Forecasting the 2024 US General Election"
subtitle: "Analyzing pollster data"
author: 
  - Sakura Hu
  - Luka Tosic
thanks: "Code and data are available at: https://github.com/LukaTosic09/Forecasting-2024-Election"
date: today
date-format: long
abstract: "This paper uses the poll of polls method to create a general linear model to predict the outcome of the 2024 US general election.
We find that current polling predicts ______ as the winner of the election. Third sentence. A limitation of the model is that it does not account for the electoral college in predicting the result, and is thus based only on the 
popular vote."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

^^^^^Sakura, I wasn't sure if I should delete the dummy code, I just kept it in for now^^^^^

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The US presidental election ...

# Estimand

Lorem Ipsum # lorem ipsum is just a placeholder in case it's unfamiliar

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Measurement

Lorem Ipsum

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Lorem Ipsum"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix 1 {-}

Lorem Ipsum

# Appendix 2 {-} #what is {-}?

Lorem Ipsum

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-10-25 paper/paper.qmd
move the files one level forward

--- 2024-10-21 starter_folder-main/paper/paper.qmd
@@
 ^^^^^Sakura, I wasn't sure if I should delete the dummy code, I just kept it in for now^^^^^
+Also, I would recommend creating a new branch locally to work in and push to, after which we can merge work into the main branch
 
@@
 
-Lorem Ipsum # lorem ipsum is just a placeholder in case it's unfamiliar
+^^^^^(discuss this)
+
+Lorem Ipsum
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+^^^^^^(I can write the whole of this section myself)
+
+Our data is the Project 538 (idk how to cite them) Presidential General Election polls (current cycle) from October 17th.
+It contains 16000 observations of polls with __ unique polls with unque poll id from __ pollsters with unique pollster id.
+The methodology of each poll was as an online panel, an app panel, a live phone call, ...
+, and/or a probability panel.
+Each poll has been given an associated pollscore, numeric grade and transparency score.
+We also know the duration of the poll by start and end date, as well as the time of publication and the number of particpants.
+The candidate sponsoring each poll is given as well as the candidate endorsed by each poll.
+Of course, each poll has a predicted outcome which is captured in the data by the
+
+-> There's a lot of other variables, but they mostly seem redundent to talk about, however we are still expected to explain them.
 
@@
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+Lorem Ipsum
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+^^^^^^(We need to discuss this)
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+We will use the poll of polls method to create a generalized linear model to create a logistic regression model to predict the outcome of the general election.
+This method allows us to get a more comprehensive estimate of election outcomes,
+
+# (from the starter code) Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+^^^^^(Discuss this as well)
+
+Our model will be of the form: logistic regression model logit(P)=log(p/1-p)=Y=Beta_0*x_1 + ... + Beta_k*x_k, p = c(x_0, x_1, ... , x_k)
+We define $Y$ the response variable as the the winner of the election, this variable come from the answer column in the raw dataset.
+
+# (from starter code) Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
 
@@
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
+Lorem Ipsum
 

### 2024-10-29 scripts/02-data_cleaning.R
cleaned data for analysis

@@ -1,44 +1,45 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Clean the election data from FiveThirtyEight for the 2024 election. 
+# Author: Tara Chakkithara
+# Date: October 29 2024
+# Contact: tara.chakkithara@icloud.com
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
-#### Workspace setup ####
+
+### Required Packages ###
 library(tidyverse)
 
-#### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+# read data 
+data <- read_csv('./data/raw_data/president_polls.csv')
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+# For our analysis we will only be looking at swing states and two of
+# the election candidates : Donald Trump and Kamala Harris. 
+swing_states <- c("Pennsylvania", "Nevada", "North Carolina", "Wisconsin",
+                  "Michigan", "Georgia", "Arizona")
+data <- data |>
+  filter(state %in% swing_states) |>
[…]

### 2024-10-30 scripts/03-test_data.R
test clean data

@@ -1,15 +1,40 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests analysis data for any issues 
+# Author: Tara Chakkithara
+# Date: October 29 2024
+# Contact: tara.chakkithara@icloud.com
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
 
-#### Workspace setup ####
+#### Required Packages ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
+# read data
+data <- read_csv('./data/analysis_data/analysis_data.csv')
 
 #### Test data ####
+# Some important things we need to check is that there are no empty entries in our data. 
+
+empty_swing_state <- data |>
+  filter(is.na(swing_state))
+
+empty_candidate_name <- data |>
+  filter(is.na(candidate_name))
+
+empty_pct <- data |>
+  filter(is.na(pct))
+
+# The following must be true 
+
+print(nrow(empty_swing_state) == 0) 
+print(nrow(empty_candidate_name) == 0)
+print(nrow(empty_pct) == 0)
+
+# Check if pct is a number between 0 and 100. 
+print(all(data$pct >= 0 & data$pct <= 100))
+
+# Check if candidate names are Kamala Harris or Donald Trump.  
+print(all(data$candidate_name %in% c("Kamala Harris", "Donald Trump")))
+
+# All print statements must be true. 
+

### 2024-10-30 scripts/04-model.R
updated model.R preamble

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: We will use bayesian forecasting with 
+# monte carlo simulation to model the 2024 elction. 
+# Author: Tara Chakkithara
+# Date: October 29 2024
+# Contact: tara.chakkithara@icloud.com
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
 
 #### Workspace setup ####

### 2024-10-30 starter_folder-main/paper/paper.qmd
completed appendix 2

@@
 date-format: long
-abstract: "This paper uses the poll of polls method to create a general linear model to predict the outcome of the 2024 US general election.
-We find that current polling predicts ______ as the winner of the election.
+abstract: "This paper uses the poll of polls method to create a general linear model to predict the outcome of the 2024 US general election.We find that current polling predicts ______ as the winner of the election.
 Third sentence.
-A limitation of the model is that it does not account for the electoral college in predicting the result, and is thus based only on the
-popular vote."
+A limitation of the model is that it does not account for the electoral college in predicting the result, and is thus based only on the popular vote."
 format: pdf
@@
 library(tidyverse)
-library(palmerpenguins)
+library(janitor)
+library(lubridate)
+library(broom)
+library(modelsummary)
+library(rstanarm)
+library(splines)
+
 ```
-
-^^^^^Sakura, I wasn't sure if I should delete the dummy code, I just kept it in for now^^^^^
 
@@
 
-# Estimand
+#Estimand
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+Our data is the Project 538 Presidential General Election polls dataset (current cycle) from October 17th.
+It contains roughly 16000 observations of polls with __ unique polls with unque poll id from __ pollsters with unique pollster id.
+The methodology of each poll was as an online panel, an app panel, a live phone call, ...
+, and/or a probability panel.
+Each poll has been given an associated pollscore, numeric grade and transparency score.
+We also know the duration of the poll by start and end date, as well as the time of publication and the number of particpants.
+The candidate sponsoring each poll is given as well as the candidate endorsed by each poll.
+Of course, each poll has a predicted outcome which is captured in the data by the
+
+Each poll has an associated pollster, and each pollster is assigned a numeric grade (0.5 to 3.0) and pollscore (-1 to 1) corresponding to the quality and reliability of the pollster and a transparency score (0 to 10) corresponding to the level of transparency the pollster provides in regards to their methodology.
+Furthemore, each poll has an associated methodology.
+
+Importantly for our modeling, ...
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| include: false
+#| warning: false
+#| message: false
 #| echo: false
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
+election_data <- read.csv("../data/raw_data/president_polls.csv")
+real_data <- election_data %>% filter(hypothetical=="false", answer == "Harris")
+
+real_data <- real_data %>% mutate(national = ifelse(is.na(state), TRUE, FALSE))
+quality_data <- real_data %>% filter(sample_size >= 1000, numeric_grade>=3.0)
+
+#still having trouble with dates
+#real_data$end_date <- as.Date(real_data$end_date, format="%d/%m/%y")
+#real_data <- real_data %>% mutate(end_date = mdy(end_date))# %>% filter(!is.na(end_date))
+#fixed_date <- ymd("2024-25-10")
+#real_data$recency <- as.numeric(difftime(fixed_date, real_data$end_date, units = "days"))
+real_data %>% mutate(end_date = mdy(end_date))
+
+```
+
+```{r}
+ggplot(real_data, aes(x=numeric_grade, y=pollscore)) + geom_point() + geom_smooth(formula = y ~ x, method = "lm",se = FALSE)+labs(title = "what's the relationship even?")
 ```
@@
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
 #| echo: false
@@
 #| message: false
+#| label: tbl-predictor vs response plots
+#| height: 20
+#| width: 6
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+ggplot(real_data, aes(x=pollster, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "pollster vs pct") #+ theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+ggplot(real_data, aes(x=state, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "state vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+
+ggplot(real_data, aes(x=methodology, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "method vs pct") #+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+
+ggplot(real_data, aes(x=methodology)) + geom_bar() + labs(title="methodology occurences") +   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+
+ggplot(real_data, aes(x=partisan, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "partisan vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+
+```
+
+```{r}
+ggplot(real_data, aes(x=sponsor_candidate_party, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "sponsor party vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+
+ggplot(real_data, aes(x=transparency_score, y=numeric_grade)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "trans vs poll") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+```{r}
+test_model <- lm(pct ~ end_date + state, data = quality_data)
+test_model
+
 ```
@@
 
-# Appendix 2 {-} #what is {-}?
+# Appendix 2 {Idealized Survey}
 
-Lorem Ipsum
+We are provided a budget of $100'000 to forecast the 2024 presidential election.
+First we must define our key terms.
+We are looking to forecast the winner of the US general election, so our key parameter of interest for survey participants is the candidate they would support and/or vote for.
+There are other secondary parameters to discuss later.
+Our target population is registered voters in the US as they would be the ones informing our parameter of interest, with our sampling frame being registered voters that we can reach with our survey.
+The sample then is the registered voters who end up taking the survey.
+We must also define a sample size that is both realistically achievable, and adequately large to make a meaningful prediction of the election winner.
+A sample size in the range of 2500-6000 is within the realm of possibility and would be sufficient.
+
+#2.1 Sampling
+
+The first task is to develop a sample from a sampling frame through some sample method(s).
+We would make use of our large budget in order to go about a probabilistic sampling strategy, specifically a stratified sampling approach with additional simple random sampling within the given strata.
+We would stratify along state, age, education, sex, income, and self-identified political affiliation.
+The motivation behind these strata is that we know these factors create bias; note that democrats are more likely to answer surveys [], as are those with higher educational attainment [].
+Additionally, certain battleground states as they are referred to like Georgia or Pennsylvania are important to gather data on as they are considered the most important states in deciding the election []; another reason is that the US population is not split evenly among states, and even in a large poll we would expect only a small, and less reliable amount of responses from certain states with smaller populations.
+
+The specifics of the stratification along the various strata will depend on facts about distribution of population, educational attainment, age, sex, and more among registered voters.
+Once such statistics are collected likely through census data, we can specify the quantities we hope to see in our sample.
+ie if we have a sample of size n, and a population of registered voters in a given state of x, we hope to see in our data (x*n)/y where y is the number of registered voters in th US.
+
+#2.2 Survey Method
+
+The survey method we will employ is an online panel of questions in the form of a Google form.
+These forms are secure and anonymous, and provide a simple way to store the results we receive.
+Additionally, the format is low cost to produce and takes no additional labour to maintain unless issues arise, in contrast with a live phone that must be either directly manned or monitored.
+Another advantage is the transparency provided, users can read and reread instructions, and return to questions they are unsure of easily.
+Keeping the form short, up to 15 minutes can help stave of disengagement from possible participants.
+
+The maintaining of the survey would take some paid labour.
+The survey would run for 1 to 2 weeks over the course of 3 months prior to the election as a way to collect data that captures shifting trends in public opinion.
+Doing this would require us to note when the survey was submitted.
+
+#2.3 Recruitment
+
+Distributing the survey will require multiple things.
+We must choose how we will distribute the survey.
+To keep people who are not registered voters from answering, there must be a way for participants to verify that they are registered voters.
+We propose they register with an email to receive the survey, and they would only receive the survey if their information matches a registered voter in the US, this system would have to be automated and not save sensitive information in order to maintain anonymity.
+Of course, we'd need to somehow access information on registered voters in the US.
+Such a system would of course take a portion of our budget to implement
+
+In order to incentivize participation, we could allocate part of the budget to providing a incentive of monetary value for participants.
+Something like a small giftcard or voucher that is guaranteed upon survey submission.
+
+To reach potential participants, a campaign to send advertise the survey would be launched.
+We could pay businesses to send a promotional email to users of their service, particularly of services used by demographics we are trying to account for in our strata.
+A powerful tool for advertising this survey would social media, especially since in the modern day, social media are used by a very wide range of people from the most partisan Republicans and Democrats to the median individual.
+We can advertise physically as well, more so in states that we are hoping to account for per our stratification.
+
+#2.4 Data Validation
+
+Once the data is collected, it would have to be parsed and cleaned.
+We would want to know how many questions were unanswered, and to produce basic summary statistics on things like the demographics and state of participants in order to gain an initial understanding of our data.
+Doing this while the survey is running can tell us if we need to  push it further in certain parts of the nation or to certain groups with the hope of amassing more responses in those categories.
+
+As survey cycles continue, we would clean the data to remove non-sensical or uninterpretable responses as well as automated or repeated responses.
+
+The next part of validation is applying stratification weights to ensure our data conforms to the stratification criteria we outlined above.
+In other words, apply post-stratification weights to align survey responses with Census-based demographics, using our previously outlined strata as variables.
+Use iterative proportional fitting (raking) to adjust for sample imbalances.
+
+#2.5 Poll Aggregation and Forecasting
+
+We would aggregate weekly poll results, adjusting for recent responses with higher weights [].
+Next we would employ Bayesian updating as in our paper to forecast trends based on historical election and polling data, which accounts for typical shifts and volatility in key regions.
+
+We would also have to account for a margin of error and integrate relevant data (e.g., voter registration numbers, early voting patterns) into final our models to refine predictions, particularly in aforementioned battleground states.
+
+#2.6 Survey Example
+
+A complete example survey can be found here: https://forms.gle/tkFYQruLWuwCMLAm9
+
+The questions provided are:
+Are you currently a registered voter?
+What is the current state you are registered to vote in?
+How would you describe your political leanings?
+Indicate your age in years
+Indicate your sex
+What is your highest level of educational attainment?
+What was your 2023 taxable income amount? (can prefer not to say)
+If the general election were held tomorrow, what party would you vote for?
+How Likely are you to vote in this election?
+If you voted in the 2020 general election, who did you vote for?
+How well do you feel your most preferred candidate's policies address your concerns and most important issues as a voter
+What are your most top 3 most important issues when deciding who to vote for?
+How confident are you in the outcome of this 2024 election?
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
-\newpage
-
 # References

### 2024-10-30 scripts/02-data_cleaning.R
completed initial model, it overfits a little

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Clean the election data from FiveThirtyEight for the 2024 election. 
+# Purpose: Clean election data from FiveThirtyEight for the 2024 election.
 # Author: Tara Chakkithara
 # Date: October 29 2024
 # Contact: tara.chakkithara@icloud.com
@@ -12,13 +12,12 @@ library(tidyverse)
 # read data 
 data <- read_csv('./data/raw_data/president_polls.csv')
 
-# For our analysis we will only be looking at swing states and two of
-# the election candidates : Donald Trump and Kamala Harris. 
+# For our analysis we will only be looking at swing states.
 swing_states <- c("Pennsylvania", "Nevada", "North Carolina", "Wisconsin",
                   "Michigan", "Georgia", "Arizona")
 data <- data |>
   filter(state %in% swing_states) |>
-  filter(candidate_name == "Kamala Harris" | candidate_name == "Donald Trump") |>
+  filter(candidate_name == "Donald Trump") |>
   rename(swing_state = state)
 
 # To ensure that we are looking at high quality polls we can filter out polls with 
@@ -32,11 +31,11 @@ data <- data |>
   mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
   filter(start_date > biden_drop_date)
   
-# Finally we can just look at the variables swing state, candidate_name, and pct
+# Finally we can just look at the variables swing state, and pct
 # for our analysis. 
 
 data <- data |>
-  select(swing_state, candidate_name, pct)
+  select(swing_state, pct)
 
 # save cleaned data for analysis in data/analysis_data
 write_csv(data, './data/analysis_data/analysis_data.csv')

### 2024-10-30 scripts/03-test_data.R

@@ -18,23 +18,17 @@ data <- read_csv('./data/analysis_data/analysis_data.csv')
 empty_swing_state <- data |>
   filter(is.na(swing_state))
 
-empty_candidate_name <- data |>
-  filter(is.na(candidate_name))
-
 empty_pct <- data |>
   filter(is.na(pct))
 
 # The following must be true 
 
 print(nrow(empty_swing_state) == 0) 
-print(nrow(empty_candidate_name) == 0)
 print(nrow(empty_pct) == 0)
 
 # Check if pct is a number between 0 and 100. 
 print(all(data$pct >= 0 & data$pct <= 100))
 
-# Check if candidate names are Kamala Harris or Donald Trump.  
-print(all(data$candidate_name %in% c("Kamala Harris", "Donald Trump")))
-
 # All print statements must be true. 
 
+#### We should test our model ####

### 2024-10-30 scripts/04-model.R

@@ -14,23 +14,85 @@ library(rstanarm)
 #### Read data ####
 analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+# phase 1. we can generate the KDEs of pct for each swing state. 
+generate_state_kde <- function(data, state){
+  filtered_data <- data |>
+    filter(swing_state == state)
+  kde <- density(filtered_data$pct)
+  kde_data <- data.frame(x = kde$x, y = kde$y)
+  return(kde_data)
+}
+
+# phase 2. we can simulate a pct value for each state. 
+simulate <- function(data, state) {
+  kde_data <- generate_state_kde(data, state)
+  random_sample <- sample(kde_data$x, size = 1, prob = kde_data$y,
+                          replace = TRUE)
+  
+  return(random_sample/100)
+}
+
+# phase 3. we can allocate seats per swing state.
+add_seats <- function(data, state, seats) {
+  prob_win <- simulate(data, state)
+  # here we use pct to generate a win or a loss. 
+  outcome <- rbinom(1, 1, prob_win)
+  return(seats * outcome)
+}
+
+# phase 4. we can simulate a single election.
+simulate_election <- function(data) {
+  # we initially assign trump all the red states and randomly allocate 
+  # seats based on pct scores. 
+  seats <- 219
+  seats <- seats + add_seats(data, "Arizona", 11)
+  seats <- seats + add_seats(data, "Nevada", 6)
+  seats <- seats + add_seats(data, "Wisconsin", 10)
+  seats <- seats + add_seats(data, "Michigan", 15)
+  seats <- seats + add_seats(data, "Pennsylvania", 19)
+  seats <- seats + add_seats(data, "North Carolina", 16)
+  seats <- seats + add_seats(data, "Georgia", 16)
+  return(seats / 538)
+}
+
+# phase 5. we can simulate an election multiple times. 
+monte_carlo_simulation <- function(data, num_simulations) {
+  results <- numeric(num_simulations)
+  
+  for(i in 1:num_simulations) {
[…]

### 2024-10-30 scripts/04-model.R
final model complete, predicts a harris win

@@ -6,15 +6,15 @@
 # Contact: tara.chakkithara@icloud.com
 # License: MIT
 
-
-#### Workspace setup ####
+### Required Packages ###
 library(tidyverse)
-library(rstanarm)
 
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+# read data
+data <- read_csv('data/analysis_data/analysis_data.csv')
 
-# phase 1. we can generate the KDEs of pct for each swing state. 
+set.seed(1)
+
+# for all swing states generate KDEs for pct
 generate_state_kde <- function(data, state){
   filtered_data <- data |>
     filter(swing_state == state)
@@ -23,7 +23,7 @@ generate_state_kde <- function(data, state){
   return(kde_data)
 }
 
-# phase 2. we can simulate a pct value for each state. 
+# simulate a pct for a state
 simulate <- function(data, state) {
   kde_data <- generate_state_kde(data, state)
   random_sample <- sample(kde_data$x, size = 1, prob = kde_data$y,
@@ -32,18 +32,16 @@ simulate <- function(data, state) {
   return(random_sample/100)
 }
 
-# phase 3. we can allocate seats per swing state.
+# generate seats based on simulated pct for a swing state
 add_seats <- function(data, state, seats) {
   prob_win <- simulate(data, state)
-  # here we use pct to generate a win or a loss. 
   outcome <- rbinom(1, 1, prob_win)
   return(seats * outcome)
 }
 
-# phase 4. we can simulate a single election.
+# simulate an election round
 simulate_election <- function(data) {
-  # we initially assign trump all the red states and randomly allocate 
-  # seats based on pct scores. 
+  # initially set seat count to include red states
   seats <- 219
   seats <- seats + add_seats(data, "Arizona", 11)
   seats <- seats + add_seats(data, "Nevada", 6)
@@ -55,7 +53,7 @@ simulate_election <- function(data) {
   return(seats / 538)
 }
 
-# phase 5. we can simulate an election multiple times. 
+# helper to simulate the election multiple times. 
 monte_carlo_simulation <- function(data, num_simulations) {
   results <- numeric(num_simulations)
   
@@ -66,14 +64,15 @@ monte_carlo_simulation <- function(data, num_simulations) {
   return(results)
 }
 
-# phase 6. using all simulations we can update our beta prior. 
[…]

### 2024-10-31 scripts/03-test_data.R
simulated analysis data

@@ -30,5 +30,3 @@ print(nrow(empty_pct) == 0)
 print(all(data$pct >= 0 & data$pct <= 100))
 
 # All print statements must be true. 
-
-#### We should test our model ####

### 2024-10-31 paper/paper.qmd

--- 2024-10-21 starter_folder-main/paper/paper.qmd
@@
 
-# Appendix 2 {-} #what is {-}?
+# Appendix 2 {Idealized Survey}
 
-Lorem Ipsum
+We are provided a budget of $100'000 to forecast the 2024 presidential election.
+First we must define our key terms.
+We are looking to forecast the winner of the US general election, so our key parameter of interest for survey participants is the candidate they would support and/or vote for.
+There are other secondary parameters to discuss later.
+Our target population is registered voters in the US as they would be the ones informing our parameter of interest, with our sampling frame being registered voters that we can reach with our survey.
+The sample then is the registered voters who end up taking the survey.
+We must also define a sample size that is both realistically achievable, and adequately large to make a meaningful prediction of the election winner.
+A sample size in the range of 2500-6000 is within the realm of possibility and would be sufficient.
+
+# 2.1 Sampling
+
+The first task is to develop a sample from a sampling frame through some sample method(s).
+We would make use of our large budget in order to go about a probabilistic sampling strategy, specifically a stratified sampling approach with additional simple random sampling within the given strata.
+We would stratify along state, age, education, sex, income, and self-identified political affiliation.
+The motivation behind these strata is that we know these factors create bias; note that democrats are more likely to answer surveys [], as are those with higher educational attainment [].
+Additionally, certain battleground states as they are referred to like Georgia or Pennsylvania are important to gather data on as they are considered the most important states in deciding the election []; another reason is that the US population is not split evenly among states, and even in a large poll we would expect only a small, and less reliable amount of responses from certain states with smaller populations.
+
+The specifics of the stratification along the various strata will depend on facts about distribution of population, educational attainment, age, sex, and more among registered voters.
+Once such statistics are collected likely through census data, we can specify the quantities we hope to see in our sample.
+ie if we have a sample of size n, and a population of registered voters in a given state of x, we hope to see in our data (x*n)/y where y is the number of registered voters in th US.
+
+# 2.2 Survey Method
+
+The survey method we will employ is an online panel of questions in the form of a Google form.
+These forms are secure and anonymous, and provide a simple way to store the results we receive.
+Additionally, the format is low cost to produce and takes no additional labour to maintain unless issues arise, in contrast with a live phone that must be either directly manned or monitored.
+Another advantage is the transparency provided, users can read and reread instructions, and return to questions they are unsure of easily.
+Keeping the form short, up to 15 minutes can help stave of disengagement from possible participants.
+
+The maintaining of the survey would take some paid labour.
+The survey would run for 1 to 2 weeks over the course of 3 months prior to the election as a way to collect data that captures shifting trends in public opinion.
+Doing this would require us to note when the survey was submitted.
+
+# 2.3 Recruitment
+
+Distributing the survey will require multiple things.
+We must choose how we will distribute the survey.
+To keep people who are not registered voters from answering, there must be a way for participants to verify that they are registered voters.
+We propose they register with an email to receive the survey, and they would only receive the survey if their information matches a registered voter in the US, this system would have to be automated and not save sensitive information in order to maintain anonymity.
+Of course, we'd need to somehow access information on registered voters in the US.
+Such a system would of course take a portion of our budget to implement
+
+In order to incentivize participation, we could allocate part of the budget to providing a incentive of monetary value for participants.
+Something like a small giftcard or voucher that is guaranteed upon survey submission.
+
+To reach potential participants, a campaign to send advertise the survey would be launched.
+We could pay businesses to send a promotional email to users of their service, particularly of services used by demographics we are trying to account for in our strata.
+A powerful tool for advertising this survey would social media, especially since in the modern day, social media are used by a very wide range of people from the most partisan Republicans and Democrats to the median individual.
+We can advertise physically as well, more so in states that we are hoping to account for per our stratification.
+
+# 2.4 Data Validation
+
+Once the data is collected, it would have to be parsed and cleaned.
+We would want to know how many questions were unanswered, and to produce basic summary statistics on things like the demographics and state of participants in order to gain an initial understanding of our data.
+Doing this while the survey is running can tell us if we need to  push it further in certain parts of the nation or to certain groups with the hope of amassing more responses in those categories.
+
+As survey cycles continue, we would clean the data to remove non-sensical or uninterpretable responses as well as automated or repeated responses.
+
+The next part of validation is applying stratification weights to ensure our data conforms to the stratification criteria we outlined above.
+In other words, apply post-stratification weights to align survey responses with Census-based demographics, using our previously outlined strata as variables.
+Use iterative proportional fitting (raking) to adjust for sample imbalances.
+
+# 2.5 Poll Aggregation and Forecasting
+
+We would aggregate weekly poll results, adjusting for recent responses with higher weights [].
+Next we would employ Bayesian updating as in our paper to forecast trends based on historical election and polling data, which accounts for typical shifts and volatility in key regions.
+
+We would also have to account for a margin of error and integrate relevant data (e.g., voter registration numbers, early voting patterns) into final our models to refine predictions, particularly in aforementioned battleground states.
+
+# 2.6 Survey Example
+
+A complete example survey can be found here: https://forms.gle/tkFYQruLWuwCMLAm9
+
+The questions provided are:
+Are you currently a registered voter?
+What is the current state you are registered to vote in?
+How would you describe your political leanings?
+Indicate your age in years
+Indicate your sex
+What is your highest level of educational attainment?
+What was your 2023 taxable income amount? (can prefer not to say)
+If the general election were held tomorrow, what party would you vote for?
+How Likely are you to vote in this election?
+If you voted in the 2020 general election, who did you vote for?
+How well do you feel your most preferred candidate's policies address your concerns and most important issues as a voter
+What are your most top 3 most important issues when deciding who to vote for?
+How confident are you in the outcome of this 2024 election?
 

### 2024-11-01 paper/paper.qmd
Update paper.qmd
add first draft of appendix I

@@
 
-Lorem Ipsum
+The pollster that is chosen to be analyzed is The New York Times/Siena College.
+The population targeted by this poll consists of the entire U.S.
+electorate.
+The frame for the poll is derived from the L2 voter file, which is a nonpartisan database, and further supplemented by matched cellular telephone numbers from Marketing Systems Group (The New York Times, 2024).
+From this frame, respondents are randomly selected from a national list of registered voters to participated in telephone interviews in English and Spanish (The New York Times, 2024).
+To enhance its regional and demographic representativeness, the national poll includes separate samples for Florida (622 voters) and Texas (617 voters) and applies oversampling to ensure adequate responses from key demographic groups, specifically Black voters and Hispanic voters (The New York Times, 2024).
+The poll’s oversample includes 589 Black voters and 902 Hispanic voters, with subcategories for individuals identifying as a single or multiple races or ethnicities (The New York Times, 2024).
+By adjusting the weights of these subgroups in the final analysis, the poll aims to produce results reflective of the national electorate (The New York Times, 2024).
+
+The sample for this poll was recruited through a stratified sampling method, which considered factors such as state, party affiliation, race, gender, and age (The New York Times, 2024).
+Each stratum was adjusted to address differential coverage and expected response rates, with the initial selection weights reflecting the reciprocal of a stratum’s telephone coverage and modeled response rate based on historical nonresponse data from prior polls (The New York Times, 2024).
+The poll’s stratification process included additional adjustments by state to accommodate varying levels of telephone number productivity (The New York Times, 2024).
+The sampling approach allows for representation of diverse demographics and achieves a more balanced sample, but it introduces trade-offs.
+While oversampling minority groups improves representation and allows for deeper analysis, the increased sampling error for these smaller subgroups complicates the interpretation of subgroup findings relative to the broader electorate (The New York Times, 2024).
+
+Nonresponse is managed through stratified sampling adjustments, where mean response rates for each stratum help inform weighting to minimize potential bias from nonresponse (The New York Times, 2024).
+Additionally, the survey package in R is used to apply a series of weights, which adjusts the data based on various demographic characteristics and compensates for nonresponse at both the sample and question level (The New York Times, 2024).
+For instance, demographic adjustments are made based on age, party affiliation, race, and educational attainment (The New York Times, 2024).
+Moreover, among respondents with multiple contact numbers, the poll selects the number with the highest predicted response rate, further mitigating nonresponse bias and increasing the likelihood of respondent participation (The New York Times, 2024).
+
+The questionnaire used in The New York Times/Siena College poll has several strengths aimed at enhancing accuracy, representativeness, and reliability.
+One advantage lies in the survey's duration, with calls limited to less than 15 minutes to maintain respondent engagement (The New York Times, 2024).
+This approach helps reduce the likelihood of incomplete responses due to participant fatigue, thereby supporting data integrity.
+By incorporating a stratified sampling approach that oversamples in battleground states, the questionnaire captures opinions from critical regions likely to influence the presidential election outcome (The New York Times, 2024).
+This focus aligns with the Electoral College system, making the data more actionable for understanding the dynamics within these decisive states.
+Moreover, the reliance on voter registration files, which contain detailed demographic and geographic data for 200 million Americans, supports robust sample balancing and demographic matching.
+The voter file's information is used to ensure proportional representation across party lines, regions, race, age, and other critical attributes.
+This allows the questionnaire to better represent national diversity and enhances accuracy through weighting adjustments for underrepresented demographics, such as those without a college degree (The New York Times, 2024).
+The use of telephone interviews, despite declining response rates, remains one of the few methods to quickly access a randomized sample of voters.
+Alternative methods like email or mail recruitment introduce their own issues, such as attracting only highly interested individuals, potentially biasing the sample toward the politically active rather than the average voter (The New York Times, 2024).
+Furthermore, the poll's careful management of sampling error, with a +- 2.4-point margin for the national sample and +- 5 points for state polls, ensures a strong level of accuracy while acknowledging the inherent uncertainty in survey results.
+
+Despite these advantages, the questionnaire also faces several limitations.
+As response rates have decreased in recent years, the representativeness of telephone surveys has likely suffered, as individuals are more reluctant to answer phone calls from unfamiliar numbers.
+This trend has increased the costs of conducting such surveys, as more calls are needed to reach a proportional sample, and some demographic groups may remain underrepresented despite weighting adjustments (The New York Times, 2024).
+The reliance on telephone interviews also risks excluding younger voters and others who rely primarily on digital communication, potentially skewing the data if not properly addressed through demographic balancing and weighting.
+Additionally, while the questionnaire's margin of error provides a statistical measure of confidence, this margin can be significant, particularly in close races where even a two-point margin could change interpretations and outcomes.
+This issue highlights the limitations of polling as a "blunt instrument" where minor fluctuations in results can feel disproportionately impactful in a close election context (The New York Times, 2024).
 

### 2024-11-03 paper/paper.qmd
added part of an introduction and data section for teammate parsing

--- 2024-10-31 paper/paper.qmd
@@
   - Luka Tosic
+  - Tara Chakkithara
 thanks: "Code and data are available at: https://github.com/LukaTosic09/Forecasting-2024-Election"
@@
 
-The US presidental election ...
+The United States has been the World's pinnacle of Hegemonic power for the last eighty to one hundered years, a nation-state powerful enough to exert its economic and military power as well as its societal and cultural influence on virtually all corners of the Globe.
+As much as many of our fellow Non-Americans loathe to hear it, the politics of the United States matter to people globally because their effect is felt globally.
+American citizens on the other hand have understand immediately the importance of the general election in influencing the direction their nation takes.
 
-#Estimand
+The domestic political situation has escalated in the US for years, with higher rates of homelessness and income inequality being major points of concern.
+Moreover, alt right movements continue to gather steam targeting women's rights, minority rights and immigration.
+The current Democrat administration continues to be lambasted for its adherence to maintaining the status quo, with many openly protesting against the US's relationship with Israel and foreign involvement generally.
+We have also seen demonstration against the stripping away of reproductive rights, and against police brutality contrasted by anti-immigration marches.
 
-Lorem Ipsum # lorem ipsum is just a placeholder in case it's unfamiliar
+The primary candidates in this election are the Democrat's Kamala Harris, and the Republican's Donald Trump.
+The estimand we will be using is a probability distribution of Trump's win percentage, which of course would inform Harris's in turn.
+Our estimate of this estimate will be found by analyzing the previous two elections, 2016's Clinton vs Trump and 2020's Trump vs Biden, as well as current polling data to make a prediction on either candidate's election rate and thus the winner of the election.
+Said estimate will allow us to make a prediction on who we think is likelier to win the election.
+
+An important aspect of all elections, but particularly this one owing to the current volatile and unpredictable political climate, is the notion of so called "Swing" or "Battleground" states.
+In a presidential election, each state is awarded a number of electoral votes based on the state's total representation in congress, or more simply, its size.
+There are a total of 538 electoral votes, and a candidate needs 270 to secure a majority and the election.
+Some states have no consistent voting pattern over elections, these are the aforementioned swing states.
+We will focus our analysis on swing states in order to make an informed prediction about them and the election as a whole.
+
+In particular, we used Bayesian updating and Monte Carlo Simulations of the election to model the results.
+We found this method of modeling was more effective than something like a Generalized Linear Model due to the nature of the variables found in the dataset.
+More information about the shape of the data is found in the data section, but in short we did not think that most of the variables present were capable creating a GLM that would accurately predict the results as none of the variables reasonably predict the hypothetical response of vote percentage based on our exploratory data analysis.
+
+Our high level results are ...
+
+This is important because ...
+
+*include estimand
 
@@
 
-Our data is the Project 538 Presidential General Election polls dataset (current cycle) from October 17th.
-It contains roughly 16000 observations of polls with __ unique polls with unque poll id from __ pollsters with unique pollster id.
-The methodology of each poll was as an online panel, an app panel, a live phone call, ...
-, and/or a probability panel.
-Each poll has been given an associated pollscore, numeric grade and transparency score.
+# Data Overview (missing graphs and tables)
+
+Our data is the Project 538 [cite] Presidential General Election polls dataset (current cycle) from October 17th.
+The data contains roughly 16000 observations of polls taken from January 2023 at the earliest to October 16th at the latest.
+Each poll has been given an associated pollscore, numeric grade and transparency score, and the methodology used to conduct it is also given.
 We also know the duration of the poll by start and end date, as well as the time of publication and the number of particpants.
 The candidate sponsoring each poll is given as well as the candidate endorsed by each poll.
-Of course, each poll has a predicted outcome which is captured in the data by the
 
+(can be deleted)
 Each poll has an associated pollster, and each pollster is assigned a numeric grade (0.5 to 3.0) and pollscore (-1 to 1) corresponding to the quality and reliability of the pollster and a transparency score (0 to 10) corresponding to the level of transparency the pollster provides in regards to their methodology.
-Furthemore, each poll has an associated methodology.
+Furthermore, each poll has an associated methodology.
+[graphics to include?, pollscore vs numeric_grade and numeric_grade vs transparency score can show strong linear relationship, providing justification to use one]
 
-Importantly for our modeling, ...
+Important to our modeling are the variables answer and pct, where answer indicates what candidate the poll predicts will receive the associated percentage (pct) of votes.
+Thus each poll is captured at least twice in the dataset, once for each polled candidate.
+The associated unique poll_id is a variable that then captures multiple observations of the same poll.
+Another important variable is Hypothetical, which notes whether a poll is between candidates that ran against each other in reality, or in a hypothetical scenario.
+Another important variable for our modeling is state, which indicates which state, if any, the poll was focused.
+[pct by state for both candidates can show that swing states are less clear]
+
+# Data Cleaning
+
+To clean this data, we first filtered out all hypothetical polls.
+Then, we filtered out any polls not focused on swing states.
+Next,  we looked only at polls answering a percentage of Trump as he appeared in the last two presidential races, and we removed all data from from before Biden dropped out to get results for Trump vs Harris.
+
+Next, we wanted to focus only on high quality polls, so we filtered out all polls with a pollscore greater than zero.
 
@@
 Talk way more about it.
-
-# Measurement
-
-Lorem Ipsum
 
@@
 Another advantage is the transparency provided, users can read and reread instructions, and return to questions they are unsure of easily.
-Keeping the form short, up to 15 minutes can help stave of disengagement from possible participants.
+Keeping the form short, up to 15 minutes can help stave off disengagement from possible participants.
 
@@
 
-In order to incentivize participation, we could allocate part of the budget to providing a incentive of monetary value for participants.
+In order to incentivize participation, we could allocate part of the budget to providing an incentive of monetary value for participants.
 Something like a small giftcard or voucher that is guaranteed upon survey submission.
 
-To reach potential participants, a campaign to send advertise the survey would be launched.
+To reach potential participants, a campaign to advertise the survey would be launched.
 We could pay businesses to send a promotional email to users of their service, particularly of services used by demographics we are trying to account for in our strata.
-A powerful tool for advertising this survey would social media, especially since in the modern day, social media are used by a very wide range of people from the most partisan Republicans and Democrats to the median individual.
-We can advertise physically as well, more so in states that we are hoping to account for per our stratification.
+A powerful tool for advertising this survey would be social media, especially since in the modern day, social media are used by a very wide range of people from the most partisan Republicans and Democrats to the median individual.
+We can advertise physically as well, more so in the states that we are hoping to account for per our stratification.
 
@@
 
-We would also have to account for a margin of error and integrate relevant data (e.g., voter registration numbers, early voting patterns) into final our models to refine predictions, particularly in aforementioned battleground states.
+We would also have to account for a margin of error and integrate relevant data (e.g., voter registration numbers, early voting patterns) into final our models to refine predictions, particularly in the aforementioned battleground states.
 

### 2024-11-03 paper/paper.qmd
first draft of the model and results section

--- 2024-11-01 paper/paper.qmd
@@
 ---
-title: "Forecasting the 2024 US General Election"
-subtitle: "Analyzing pollster data"
+title: "The Coin Flip Effect"
+subtitle: "How Swing States Create Election Outcomes That Resemble The Likelihood of a Coin Flip"
 author: 
+  - Tara Chakkithara
   - Sakura Hu
@@
 date-format: long
-abstract: "This paper uses the poll of polls method to create a general linear model to predict the outcome of the 2024 US general election.We find that current polling predicts ______ as the winner of the election.
-Third sentence.
-A limitation of the model is that it does not account for the electoral college in predicting the result, and is thus based only on the popular vote."
+abstract: "This paper employs the poll of polls method to develop a bayesian hierarcical model to predict the outcome of the 2024 US general election.
+Our analysis, based on polling data up to October 19, 20204, indicates that Kamala Harris is predicted to win the election.
+Modeling election outcomes in the United States poses challenges due to the complexities of the electoral college system, this study specifically focuses on the electoral college vote rather than the popular vote."
 format: pdf
@@
 # Model
+## Model Justification
+The unique structure of the electoral college voting system in the United States makes it unlikely for any candidate other than Donald Trump or Kamala Harris to be competitive in the 2024 presidential election.
+Given Trump's historical performance in previous campaigns, particularly the 2016 election, it is more insightful to model his election outcomes.
+To achieve this, we have decided to create a Bayesian Hierarchical Model.
 
-The goal of our modelling strategy is twofold. Firstly,...
+Our goal is to model the probability distribution of the proportion of electoral college seats that Donald Trump may secure, utilizing @citeR.
+The 2024 election closely resembles the 2016 election, which Donald Trump won, and he is facing a female opponent again.
+To capture our initial beliefs about the election being competitive, with Trump holding a slight advantage, we have selected a weak prior of Beta(4,3).
+This prior distribution is characterized by a mode that aligns closely with Trump’s previous percentage of electoral seats allocated in 2016.
+The Beta distribution is particularly effective in Bayesian modeling due to its ease of updating with new data.
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+To refine our prior and incorporate more recent data, we will focus on the election outcomes from key swing states.
+The swing states for the 2024 general election are: Nevada, Arizona, Wisconsin, Michigan, Pennsylvania, North Carolina, and Georgia.
+Model validation and additional details are found in [Appendix -@sec-model-details].
 
-## Model set-up
+## Model Set Up
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+Let $X$ be the random variable: proportion of electoral college seats Donald Trump will win.
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+$$
+X \sim \text{Beta}(4, 3)
+$$
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+For each swing state $j$, let $Y_{j, 1}, Y_{j, 2}, Y_{j, 3}, ..., Y_{j, n_j}$ be independent and identically distributed (i.i.d) random variables, where $1 \leq j \leq 7$ and  $Y_{j, i}$ represents the percentage of votes for Donald Trump in state $j$.
+$F_{j}$ is some unknown probability distribution of voting percentages of state $j$.
 
-### Model justification
+$$
+Y_{j, i} \sim F_{j}(x) \quad \text{for } j = 1, 2, \ldots, 7 \text{ and } i = 1, 2, \ldots, n_j
+$$
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+We can use Kernel Density Estimation to estimate $F_j$.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+$$
+\hat{F}_{j}(x) = \frac{1}{n} \sum_{i=1}^{n_j} K_h(x - Y_{j, i})
+$$
+
+Where $K_h$ is the gaussian kernel with bandwidth $h = n_j − \frac{1}{5} n_j^{-\frac{1}{5}}\frac{n_j - 1}{5}$. 
+Let $\hat Z_{1}, \hat Z_{2}, ..., \hat Z_{7}$ be simulated values drawn from $\hat F_{1}, \hat F_{2}, ..., \hat F_{7}$ respectively.
+Let $M_{j}$ be the election outcome of the state (win or lose).
+$M_{j}$ can be modeled using a binomial distribution where $n_j$ is the number of polls.
+
+$$
+M_{j} \sim \text{Binomial}(n_j, \hat Z_{j})
+$$
+
+We can now simulate an election and create an estimator for $X$, the random variable representing the proportion of electoral seats Donald Trump will hold after the election.
+We start by allocating Trump all the seats of prodominatly red states which are 219 seats.
+We then add additional seats for each swing state before diving by the total number of seats to estimate the proportion of seats Trump will win.
+Here, $s_j$ is the number of electoral college seats in state $j$.
+$$
+\hat X = \frac{(s_1)\hat M_{1} + (s_2)\hat M_{2} + \ldots + (s_7)\hat M_{7} + 219}{538} 
+$$
+
+Using monte carlo simulation we can generate 10,000 estimates $\hat X$. We apply a learning rate of 0.01 to avoid overfitting and update our prior.
+
+$$
+X \sim Beta(4 + W, 3 + L)
+$$
+
+Where W is the number of estimates greater than 0.5 multiplied by the learning rate of 0.01, and L is the number of estimates less than or equal to 0.5 multiplied by the same learning rate of 0.01.
+
+## Model Assumptions & Limitations
+
+Our model relies on polling data collected after Joe Biden, the former primary candidate for the Democratic Party, withdrew from the race.
+While this data is recent, it does not account for any shifts in voter sentiment that may have occurred between October 19, 2024, and November 5, 2024.
+
+In today’s highly polarized political climate, there is an increased likelihood that voters may not respond honestly in polls.
+This lack of transparency can lead to significant data distortions, ultimately compromising the accuracy of predictions.
+
+Additionally, the model’s simplicity poses a limitation, as it does not consider various critical factors that can influence electoral outcomes.
+Economic conditions, social dynamics, and demographic shifts are all significant elements that can impact voter behavior but are not incorporated into the model.
+
+This model designed specifically for the context of United States elections.
+It is built on assumptions inherent to the electoral college system, such as the presence of two dominant candidates and a winner-takes-all allocation of electoral seats.
+As a result, the insights derived from this model may not be easily generalizable to elections in different countries or political contexts.
 
 # Results
-
-Our results are summarized in @tbl-modelresults.
 
@@
 #| message: false
+library(tidyverse)
+library(scales)
 
-library(rstanarm)
+model <- readRDS(file = here::here("models/model.rds"))
+data <- model$plot_data
+prior <- model$prior
+updated <- model$updated
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
+lower_bound <- qbeta(0.025, updated$alpha, updated$beta)
+upper_bound <- qbeta(0.975, updated$alpha, updated$beta)
+
+mode_prior <- data$x[which.max(data$density[data$type == "Prior"])]
+mode_updated <- data$x[which.max(data$density[data$type == "Updated"])]
+
+ggplot(data, aes(x = x, y = density, color = type)) +
+    geom_line(size = 1) + 
+    labs(title = "Probability Distribution of Donald Trump's Electoral College Seat Percentage",
+         x = "Proportion",
+         y = "Density") +
+    theme_minimal(base_size = 10) + 
+    scale_color_manual(values = c("forestgreen", "purple")) + 
+    theme(
+        plot.title = element_text(hjust = 0.5, size = 10),
+        legend.title = element_blank(), 
+        legend.position = "top",
+        panel.grid.major = element_line(color = "grey80"),  
+        panel.grid.minor = element_blank(),
+        axis.ticks = element_line(color = "black"), 
+        axis.ticks.length = unit(0.25, "cm"),
+        axis.title.x = element_text(margin = margin(t = 10)), 
+        axis.title.y = element_text(margin = margin(r = 10))  
+    ) +
+
+    geom_ribbon(data = filter(data, type == "Prior"), 
+                aes(ymin = 0, ymax = density), 
+                fill = "forestgreen", alpha = 0.1) +  
+    geom_ribbon(data = filter(data, type == "Updated"), 
+                aes(ymin = 0, ymax = density), 
+                fill = "purple", alpha = 0.1) + 
+    geom_segment(aes(x = mode_prior, xend = mode_prior, y = 0, 
+                     yend = max(data$density[data$type == "Prior"]) * 0.8),
+                 linetype = "dashed", color = "forestgreen", size = 1) +
+    geom_segment(aes(x = mode_updated, xend = mode_updated, y = 0, 
+                     yend = max(data$density[data$type == "Updated"]) * 0.8),
+                 linetype = "dashed", color = "purple", size = 1) +
+    scale_x_continuous(breaks = c(0, 0.23, 0.44, 0.5, 0.6, 0.68, 0.75, 1), 
+                       labels = function(x) sprintf("%.2f", x)) + 
+    geom_vline(xintercept = lower_bound, linetype = "dotted", color = "purple") +
+    geom_vline(xintercept = upper_bound, linetype = "dotted", color = "purple") +
+    annotate("text", x = lower_bound, y = max(data$density[data$type == "Updated"]) * 0.85,
+             label = "",
+             color = "purple", vjust = -1) +
+    annotate("text", x = upper_bound, y = max(data$density[data$type == "Updated"]) * 0.85,
+             label = "",
+             color = "purple", vjust = -1)
+
 ```
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Lorem Ipsum"
-#| warning: false
+The updated probability distribution for Donald Trump's electoral college seat percentage is modeled as Beta(7.67,9.33).
+This distribution has a mode of approximately 0.44, indicating that the most likely outcome is that Trump would secure around 44% of the electoral college seats.
+This suggests that Kamala Harris is favored to win the overall election.
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
+However, it’s crucial to emphasize that we cannot confidently assert that Kamala Harris will win with 95% certainty.
+The shape of the probability distribution is similar to that of a coin flip, reflecting significant uncertainty in electoral outcomes.
+While Harris may have a higher likelihood of winning, there remains a considerable chance for Trump to secure a majority of the seats.
+
+To quantify Trump’s chances of victory, we evaluated the cumulative distribution function (CDF) of the Beta distribution at the 50% threshold, which corresponds to the majority of electoral college seats.
+By subtracting this CDF value from 1, we calculated the probability of Trump winning the election.
+The result of this analysis indicates that Trump has a 35% probability of securing victory.
 

### 2024-11-03 scripts/04-model.R

@@ -91,6 +91,7 @@ update_beta_prior <- function(prior_alpha, prior_beta, data, num_simulations){
   saveRDS(result, './models/model.rds')
 }
 
-# since Donald Trump is running against a woman again 
-# our prior mode is based on 2016 election results. run 10000 simulated elections. 
+# since Donald Trump is running against a woman again.
+# our prior mode is based on 2016 election results.
+# run 10000 simulated elections. 
 update_beta_prior(4, 3, data, 10000)

### 2024-11-03 paper/paper.qmd
refined the abstract

@@
 title: "The Coin Flip Effect"
-subtitle: "How Swing States Create Election Outcomes That Resemble The Likelihood of a Coin Flip"
+subtitle: "How Swing States Transform Elections Into Games of Chance"
 author: 
@@
   - Luka Tosic
-  - Tara Chakkithara
 thanks: "Code and data are available at: https://github.com/LukaTosic09/Forecasting-2024-Election"
@@
 date-format: long
-abstract: "This paper employs the poll of polls method to develop a bayesian hierarcical model to predict the outcome of the 2024 US general election.
-Our analysis, based on polling data up to October 19, 20204, indicates that Kamala Harris is predicted to win the election.
-Modeling election outcomes in the United States poses challenges due to the complexities of the electoral college system, this study specifically focuses on the electoral college vote rather than the popular vote."
+abstract: "This paper uses the poll of polls method to develop a bayesian hierarcical model to predict the outcome of the 2024 U.S.
+election.
+Based on polling data collected up to October 19, 2024, our analysis forecasts that Kamala Harris is likely to win the election.
+Given the complexities of the electoral college system, this study specifically focuses on predicting the electoral college vote rather than the popular vote."
 format: pdf

### 2024-11-04 paper/paper.qmd

@@
 ---
-title: "The Coin Flip Effect"
-subtitle: "How Swing States Transform Elections Into Games of Chance"
+title: "The Influence of Swing States on Election Outcomes"
 author: 
@@
 date-format: long
-abstract: "This paper uses the poll of polls method to develop a bayesian hierarcical model to predict the outcome of the 2024 U.S.
+abstract: "This paper employs a Bayesian hierarchical model to forecast the outcome of the 2024 U.S.
 election.
-Based on polling data collected up to October 19, 2024, our analysis forecasts that Kamala Harris is likely to win the election.
-Given the complexities of the electoral college system, this study specifically focuses on predicting the electoral college vote rather than the popular vote."
+Based on polling data collected from July 21 to October 19, 2024, our analysis indicates that Kamala Harris is the candidate most likely to win.
+Given the complexities of the Electoral College voting system and significant political polarization, election outcomes appear very close this year.
+This paper aims to quantify this closeness and provide insights into voter behavior."
 format: pdf
@@
 library(splines)
-
+library(ggthemes)
+library(sf)
+library(dplyr)
+library(readr)
 ```
@@
 # Data {#sec-data}
+## Data Acquisition
+Data for this paper is sourced from the Project 538 Presidential General Election polls dataset @cite538.
+This dataset includes approximately 16,000 observations of polls conducted from January 2023 to October 16, 2024.
+Each poll in the dataset has an associated pollscore, sample size, start date, pct (percentage of votes), candidate name, and state.
+There are other features include time of publication, candidate sponsorship, etc, however for this analysis we will focus on the key veriables highlighted.
 
-# Data Overview (missing graphs and tables)
+## Dataset Features
+### Pollscore
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+
+election_data <- read.csv("../data/raw_data/president_polls.csv")
+
+real_data <- election_data %>% select(pollscore)
+
+ggplot(real_data, aes(x = pollscore)) +
+  geom_density(fill = "#4B8BBE", alpha = 0.6) + 
+  labs(title = "Pollscore Distribution",
+       x = "Pollscore",
+       y = "Density") +
+  theme_minimal() + 
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"), 
+    axis.title.x = element_text(size = 10, face = "plain"),             
+    axis.title.y = element_text(size = 10, face = "plain")               
+  )
+```
+
+### Sample Size
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+data <- read_csv('../data/raw_data/president_polls.csv')
+
+data <- data |>
+  select(pollscore, transparency_score)
+
+data <- data |>
+  group_by(pollscore) |>
+  summarise(avg_transparency_score = mean(transparency_score, na.rm = TRUE), .groups = "drop")
+
+ggplot(data, aes(x = pollscore, y = avg_transparency_score)) +
+  geom_point(color = 'saddlebrown', size = 3) +
+  labs(
+    x = 'Pollscore',
+    y = 'Average Transparency Score',
+    title = 'Average Transparency Score by Pollscore'
+  ) +
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 10), 
+    axis.title.x = element_text(hjust = 0.46, size = 10, margin = margin(t = 10)),  
+    axis.title.y = element_text(size = 10, margin = margin(r = 10)),
+    axis.text = element_text(size = 9)
+  )
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+election_data <- read.csv("../data/raw_data/president_polls.csv")
+
+real_data <- election_data %>% select(sample_size)
+
+ggplot(real_data, aes(x = sample_size)) +
+  geom_density(fill = "#4B8BBE", alpha = 0.6) + 
+  labs(title = "Sample Size",
+       x = "Sample Size",
+       y = "Density") +
+  xlim(0, 10000) +
+  theme_minimal() + 
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"), 
+    axis.title.x = element_text(size = 10, face = "plain"),             
+    axis.title.y = element_text(size = 10, face = "plain")               
+  )
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
+                  'North Carolina', 'Pennsylvania', 'Wisconsin')
+
+map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)
+
+data <- read_csv('../data/raw_data/president_polls.csv')
+
+biden_drop_date <- as.Date("2024-07-21")
+
+data <- data |>
+  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
+  filter(start_date < biden_drop_date) |>
+  filter(pollscore < -1) |>
+  select(state, pct, candidate_name) |>
+  filter(candidate_name %in% c("Donald Trump", "Joe Biden")) |>  
+  filter(state %in% swing_states) |>
+  group_by(state, candidate_name) |>  
+  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")
+
+map_data <- map_data |> 
+  rename(state = NAME)
+
+wide_data <- data |>
+  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
+  rename(Trump = `Donald Trump`, Biden = `Joe Biden`)
+
+merged_data <- map_data |>
+  left_join(wide_data, by = "state")
+
+merged_data <- merged_data |>
+  mutate(color_category = case_when(
+    Trump > Biden ~ "Donald Trump",
+    Biden > Trump ~ "Joe Biden",
+    TRUE ~ "Tie" 
+  ))
+
+ggplot(data = merged_data) +
+  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
+  scale_fill_manual(values = c("Donald Trump" = "red", "Joe Biden" = "blue", "Tie" = "gray"), 
+                    name = "Winner") + 
+  labs(
+    title = "Voter Sentiment During Biden's Presidential Campaign",
+    x = NULL,
+    y = NULL
+  ) +
+  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
+    legend.direction = "horizontal", 
+    legend.title = element_text(hjust = 0.5)
+  )
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
+                  'North Carolina', 'Pennsylvania', 'Wisconsin')
+
+map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)
+
+data <- read_csv('../data/raw_data/president_polls.csv')
+
+biden_drop_date <- as.Date("2024-07-21")
+
+data <- data |>
+  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
+  filter(start_date > biden_drop_date) |>
+  filter(pollscore < -1) |>
+  select(state, pct, candidate_name) |>
+  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) |>  
+  filter(state %in% swing_states) |>
+  group_by(state, candidate_name) |>  
+  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")
+
+map_data <- map_data |> 
+  rename(state = NAME)
+
+wide_data <- data |>
+  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
+  rename(Trump = `Donald Trump`, Harris = `Kamala Harris`)
+
+merged_data <- map_data |>
+  left_join(wide_data, by = "state")
+
+merged_data <- merged_data |>
+  mutate(color_category = case_when(
+    Trump > Harris ~ "Donald Trump",
+    Harris > Trump ~ "Kamala Harris",
+    TRUE ~ "Tie" 
+  ))
+
+ggplot(data = merged_data) +
+  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
+  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "Tie" = "gray"), 
+                    name = "Winner") + 
+  labs(
+    title = "Voter Sentiment after Biden Drops out of the Presidential Race",
+    x = NULL,
+    y = NULL
+  ) +
+  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
+    legend.direction = "horizontal", 
+    legend.title = element_text(hjust = 0.5)
+  )
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+
+data <- read_csv('../data/raw_data/president_polls.csv')
+
+biden_drop_date <- as.Date("2024-07-21")
+
+data <- data |>
+  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
+  mutate(end_date = as.Date(sub("(..)$", "20\\1", end_date), format = "%m/%d/%Y")) |>
+  filter(start_date > biden_drop_date) |>
+  filter(is.na(state)) |>
+  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) |>  
+  group_by(end_date, candidate_name) |>  
+  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")
+
+wide_data <- data |> 
+  pivot_wider(names_from = candidate_name, values_from = avg_pct, 
+              values_fill = list(avg_pct = NA))
+
+ggplot(data = wide_data, aes(x = end_date)) +
+  geom_line(aes(y = `Kamala Harris`, color = "Kamala Harris"), size = 1) + 
+  geom_line(aes(y = `Donald Trump`, color = "Donald Trump"), size = 1) + 
+  labs(
+    title = "Time Series of Average Polling Percentages",
+    x = "Date",
+    y = "Average Percentage",
+    color = "Candidate"
+  ) +
+  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5),
+    legend.position = "top"
+  )
+
+```
+
+# Data Overview
 
@@
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-#| echo: false
-
-election_data <- read.csv("../data/raw_data/president_polls.csv")
-real_data <- election_data %>% filter(hypothetical=="false", answer == "Harris")
-
-real_data <- real_data %>% mutate(national = ifelse(is.na(state), TRUE, FALSE))
-quality_data <- real_data %>% filter(sample_size >= 1000, numeric_grade>=3.0)
-
-#still having trouble with dates
-#real_data$end_date <- as.Date(real_data$end_date, format="%d/%m/%y")
-#real_data <- real_data %>% mutate(end_date = mdy(end_date))# %>% filter(!is.na(end_date))
-#fixed_date <- ymd("2024-25-10")
-#real_data$recency <- as.numeric(difftime(fixed_date, real_data$end_date, units = "days"))
-real_data %>% mutate(end_date = mdy(end_date))
-
-```
-
-```{r}
-ggplot(real_data, aes(x=numeric_grade, y=pollscore)) + geom_point() + geom_smooth(formula = y ~ x, method = "lm",se = FALSE)+labs(title = "what's the relationship even?")
-```
-
 Talk more about it.
@@
 (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: tbl-predictor vs response plots
-#| height: 20
-#| width: 6
-
-ggplot(real_data, aes(x=pollster, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "pollster vs pct") #+ theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-
-ggplot(real_data, aes(x=state, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "state vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-
-ggplot(real_data, aes(x=methodology, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "method vs pct") #+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-
-ggplot(real_data, aes(x=methodology)) + geom_bar() + labs(title="methodology occurences") +   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-
-ggplot(real_data, aes(x=partisan, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "partisan vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-
-```
-
-```{r}
-ggplot(real_data, aes(x=sponsor_candidate_party, y=pct)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "sponsor party vs pct") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-
-ggplot(real_data, aes(x=transparency_score, y=numeric_grade)) + geom_point() + geom_smooth(formula = y ~ x , method = "lm", color = "black", se = FALSE) + labs(title = "trans vs poll") +  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
-```
-
-```{r}
-test_model <- lm(pct ~ end_date + state, data = quality_data)
-test_model
-
-```
 

### 2024-11-04 paper/paper.qmd
Update title, abstract and intro in paper.qmd

@@
 ---
-title: "The Influence of Swing States on Election Outcomes"
+title: "Forecasting the 2024 US General Election"
+subtitle: "Analyzing the influence state on election outcome by using 538 Pollster Data with Bayesian hierarchical model, Predicting Harris Victory[need more details in result]
 author: 
@@
 date-format: long
-abstract: "This paper employs a Bayesian hierarchical model to forecast the outcome of the 2024 U.S.
-election.
-Based on polling data collected from July 21 to October 19, 2024, our analysis indicates that Kamala Harris is the candidate most likely to win.
-Given the complexities of the Electoral College voting system and significant political polarization, election outcomes appear very close this year.
-This paper aims to quantify this closeness and provide insights into voter behavior."
+abstract: "This paper forecasts the outcome of the 2024 U.S.
+general election.
+Using the poll of polls approach, a generalized linear model was developed with data from FiveThirtyEight [more spcific on method, and what is the date cutoff of the data]  The model's prediction is based on the popular vote in the seven key swing states—Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner.
+Results indicate that Harris is predicted to win, by [how much?
+need to be more spcific on the result].
+This suggests that current polling trends may provide reliable insights into the likely election result, though future shifts in voter sentiment could impact the final outcome."
 format: pdf
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+The 2024 United States presidential election, the 60th quadrennial election, is scheduled for November 5, 2024.
+The stakes are high, as the outcome will shape domestic policies and impact international relations, with effects on trade, climate change, and security.
+In recent years, the domestic political landscape has changed significantly, marked by rising homelessness, income inequality, and social unrest.
+Alt-right movements increasingly target women's rights, minority rights, and immigration policies, leading to deep societal divisions.
+The current Democratic administration faces criticism for not addressing these pressing issues, with widespread protests against U.S.
+foreign involvement and the erosion of reproductive rights.
+Protests around police brutality and anti-immigration sentiments have also surfaced, reflecting the polarized nature of contemporary American society.
 
-The United States has been the World's pinnacle of Hegemonic power for the last eighty to one hundered years, a nation-state powerful enough to exert its economic and military power as well as its societal and cultural influence on virtually all corners of the Globe.
-As much as many of our fellow Non-Americans loathe to hear it, the politics of the United States matter to people globally because their effect is felt globally.
-American citizens on the other hand have understand immediately the importance of the general election in influencing the direction their nation takes.
+The primary candidates in this election, Democratic incumbent Kamala Harris and former Republican president Donald Trump, are expected to engage in a tight race, with polling indicating that the results could be very close.
+Swing states—those that do not consistently vote for one party—are vital in determining the overall outcome, as their electoral votes can sway the final result.
+Seven important swing states—Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina—are expected to be battlegrounds in this election, making their polling data essential for accurate forecasts.
 
-The domestic political situation has escalated in the US for years, with higher rates of homelessness and income inequality being major points of concern.
-Moreover, alt right movements continue to gather steam targeting women's rights, minority rights and immigration.
-The current Democrat administration continues to be lambasted for its adherence to maintaining the status quo, with many openly protesting against the US's relationship with Israel and foreign involvement generally.
-We have also seen demonstration against the stripping away of reproductive rights, and against police brutality contrasted by anti-immigration marches.
+This paper focuses on forecasting the election outcome by employing the "poll of polls" method, utilizing data from FiveThirtyEight, along with historical data from the 2016 Clinton vs.
+Trump election and the 2020 Trump vs.
+Biden election.
+A Bayesian updating and Monte Carlo Simulations is created to predict the winner of the election outcome based on the popular vote across these swing states[some more details for what was done?].
+The primary estimand in this analysis is the probability distribution of Harris's win percentage, derived from current polling data and historical election results.
+The model predicted that Harris will win, and [add details to result]
 
-The primary candidates in this election are the Democrat's Kamala Harris, and the Republican's Donald Trump.
-The estimand we will be using is a probability distribution of Trump's win percentage, which of course would inform Harris's in turn.
-Our estimate of this estimate will be found by analyzing the previous two elections, 2016's Clinton vs Trump and 2020's Trump vs Biden, as well as current polling data to make a prediction on either candidate's election rate and thus the winner of the election.
-Said estimate will allow us to make a prediction on who we think is likelier to win the election.
+Accurate forecasting of election outcomes is vital for several reasons.
+It enhances democratic engagement by informing voters about potential results, which encourages active participation in the electoral process.
+Furthermore, political campaigns can leverage these forecasts to tailor their strategies, optimize resource allocation, and concentrate efforts on pivotal swing states that may determine the election's outcome.
+Analyzing polling data also reveals trends in voter sentiment, providing valuable insights into the electorate's priorities and concerns that can inform policy discussions.
+Ultimately, the findings from this study can improve the accuracy of predictions in a political landscape characterized by rapid change and polarization, thereby enhancing the relevance and reliability of electoral analyses.
 
-An important aspect of all elections, but particularly this one owing to the current volatile and unpredictable political climate, is the notion of so called "Swing" or "Battleground" states.
-In a presidential election, each state is awarded a number of electoral votes based on the state's total representation in congress, or more simply, its size.
-There are a total of 538 electoral votes, and a candidate needs 270 to secure a majority and the election.
-Some states have no consistent voting pattern over elections, these are the aforementioned swing states.
-We will focus our analysis on swing states in order to make an informed prediction about them and the election as a whole.
-
-In particular, we used Bayesian updating and Monte Carlo Simulations of the election to model the results.
-We found this method of modeling was more effective than something like a Generalized Linear Model due to the nature of the variables found in the dataset.
-More information about the shape of the data is found in the data section, but in short we did not think that most of the variables present were capable creating a GLM that would accurately predict the results as none of the variables reasonably predict the hypothetical response of vote percentage based on our exploratory data analysis.
-
-Our high level results are ...
-
-This is important because ...
-
-*include estimand
+The remainder of this paper is structured as follows: .......
 
 # Data {#sec-data}
-## Data Acquisition
+## Data Acquisition {#sec-data-overview}
 Data for this paper is sourced from the Project 538 Presidential General Election polls dataset @cite538.
@@
 
-## Dataset Features
+## Dataset Features {#sec-Dataset-Features}
 ### Pollscore

### 2024-11-04 paper/paper.qmd
write up on additional models

--- 2024-11-04 paper/paper.qmd
@@
 
-# Additional data details
+# Additional Data Details
 
-# Model details {#sec-model-details}
+# Model Details {#sec-model-details}
 
-## Posterior predictive check
+## Alternate Models
+In our analysis, we considered two additional models: a similar Bayesian hierarchical model with a different prior and a logistic regression model.
+
+We opted not to select logistic regression for several key reasons.
+First, generalized linear models (GLMs) provide point estimates or binary outcomes, whereas our objective was to capture a full probability distribution of the election outcomes, in addition to identifying a winner.
+Second, we observed that the percentage of votes (PCT) did not exhibit a linear relationship with any of its potential predictors.
+While this issue could potentially be addressed using splines or polynomial regression, we believed it would be better to use another modeling approach since we wanted to incorporate prior data as well.
+Due to this, we decided to adopt a Bayesian framework.
+
+The first Bayesian hierarchical model we considered employed a prior of Beta(2, 3) to model the proportion of seats that Donald Trump would win.
+This model's mode closely reflects his his 2020 election results.
+However, we decided against this prior because we felt that the 2016 election more accurately represented the current circumstances.
+Notably, even with this model, we arrived at similar conclusions: Kamala Harris was still projected to win.
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+params <- data.frame(
+  shape1 = c(2, 5.67),
+  shape2 = c(3, 9.33),
+  label = c("Prior: Beta(2, 3)", "Updated: Beta(5.67, 9.33)")
+)
+
+x_values <- seq(0, 1, length.out = 100)
+
+density_data <- do.call(rbind, lapply(1:nrow(params), function(i) {
+  data.frame(
+    x = x_values,
+    density = dbeta(x_values, params$shape1[i], params$shape2[i]),
+    label = params$label[i]
+  )
+}))
+
+ggplot(density_data, aes(x = x, y = density, color = label)) +
+  geom_line(size = 1) +
+  labs(
+    title = "Distribution of Donald Trump's Electoral College Seat Proportion",
+    x = "Probability",
+    y = "Density",
+    color = "Distribution"
+  ) +
+  theme_minimal() +
+  scale_color_manual(values = c("Prior: Beta(2, 3)" = "blue", "Updated: Beta(5.67, 9.33)" = "red")) +
+  theme(
+    plot.title = element_text(size = 10),
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    legend.title = element_text(size = 10),
+    legend.text = element_text(size = 10)
+  )
+```
+
+## Posterior Predictive Check
 

### 2024-11-04 paper/paper.qmd
update paper rendering

@@
 title: "Forecasting the 2024 US General Election"
-subtitle: "Analyzing the influence state on election outcome by using 538 Pollster Data with Bayesian hierarchical model, Predicting Harris Victory[need more details in result]
+subtitle: "Analyzing the influence state on election outcome by using 538 Pollster Data with Bayesian hierarchical model, Predicting Harris Victory[need more details in result]"
 author: 
@@
   - Luka Tosic
-thanks: "Code and data are available at: https://github.com/LukaTosic09/Forecasting-2024-Election"
+thanks: "Code and data are available at https://github.com/LukaTosic09/Forecasting-2024-Election"
 date: today

### 2024-11-04 paper/paper.qmd
updated pollscore section

@@
 ## Data Acquisition {#sec-data-overview}
-Data for this paper is sourced from the Project 538 Presidential General Election polls dataset @cite538.
+Data for this paper is sourced from the Presidential General Election polls dataset, @cite538.
 This dataset includes approximately 16,000 observations of polls conducted from January 2023 to October 16, 2024.
-Each poll in the dataset has an associated pollscore, sample size, start date, pct (percentage of votes), candidate name, and state.
-There are other features include time of publication, candidate sponsorship, etc, however for this analysis we will focus on the key veriables highlighted.
+Each poll in the dataset has an associated pollscore, sample size, start date and end data, pct (percentage of votes), candidate name, and state.
+There are other features in the dataset such as time of publication, candidate sponsorship, etc.
+For this analysis, we will focus on the key veriables mentioned above.
 
@@
 ### Pollscore
+The pollscore feature serves as a metric for evaluating the quality of a poll.
+It is a mesaurement that tries to quantify the realiablity of a poll.
+It is calculated based on factors such as poll transparency, bias, and error, @citeABC.
+A more negative pollscore indicates higher quality, reflecting a greater confidence in the poll's accuracy.
+As seen in @fig-pollscore, most polls collected by FiveThirtyEight have a pollscore under 0.
+In @fig-transparency, we see that lower poll scores are correlated with high transparency.
 
-```{r}
+```{r, fig.width=5, fig.height=2.5}
 #| warning: false
@@
 #| echo: false
-
+#| label: fig-pollscore
+#| fig-cap: "Distribution of Pollscores in the 2024 General US Election"
 election_data <- read.csv("../data/raw_data/president_polls.csv")
@@
   geom_density(fill = "#4B8BBE", alpha = 0.6) + 
-  labs(title = "Pollscore Distribution",
+  labs(title = "",
        x = "Pollscore",
@@
 
-### Sample Size
-
-```{r}
+```{r, fig.width=5, fig.height=2.5}
 #| echo: false
@@
 #| message: false
-
+#| label: fig-transparency
+#| fig-cap: "Average Transparency Scores Across Varying Pollscores"
 data <- read_csv('../data/raw_data/president_polls.csv')
@@
     y = 'Average Transparency Score',
-    title = 'Average Transparency Score by Pollscore'
+    title = ''
   ) +
@@
 
+### Sample Size
 ```{r}

### 2024-11-04 paper/paper.qmd

@@
 number-sections: true
+toc: true
 bibliography: references.bib
@@
 ### Sample Size
+The sample size of a poll is the number of people who participated in the poll.
+Generally, a larger sample size is seen as a better representation of the population.
+The most common sample size of polls is around 1250, as seen in @fig-sample-size.
+
 ```{r}
@@
 #| message: false
-
+#| label: fig-sample-size
+#| fig-cap: "Distribution of Poll Sample Sizes"
 election_data <- read.csv("../data/raw_data/president_polls.csv")
@@
   geom_density(fill = "#4B8BBE", alpha = 0.6) + 
-  labs(title = "Sample Size",
+  labs(title = "",
        x = "Sample Size",
@@
 
+### Percentage of Votes
+Each poll aims to .
+By examining how these values fluctuate over time and in response to different candidates, we can better understand their impact on public opinion.
+Currently, national level polls show Kamala Harris with a 46% approval rating compared to Donald Trump's 43%.
+This highlights the importance of candidate dynamics in shaping voter sentiment across the electorate.
+
+#### Candidates
 ```{r}
@@
 #| message: false
+library(sf)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(readr)
 
+# Define swing states
 swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
@@
 
+# Load map data
 map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)
 
+# Load polling data
 data <- read_csv('../data/raw_data/president_polls.csv')
 
+# Define Biden's withdrawal date
 biden_drop_date <- as.Date("2024-07-21")
 
-data <- data |>
-  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
-  filter(start_date < biden_drop_date) |>
-  filter(pollscore < -1) |>
-  select(state, pct, candidate_name) |>
-  filter(candidate_name %in% c("Donald Trump", "Joe Biden")) |>  
-  filter(state %in% swing_states) |>
-  group_by(state, candidate_name) |>  
+# Process data before Biden's withdrawal
+data1 <- data %>%
+  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) %>%
+  filter(start_date < biden_drop_date, pollscore < -1) %>%
+  filter(candidate_name %in% c("Donald Trump", "Joe Biden"), state %in% swing_states) %>%
+  group_by(state, candidate_name) %>%
   summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")
 
-map_data <- map_data |> 
+# Prepare map data
+map_data <- map_data %>%
   rename(state = NAME)
 
-wide_data <- data |>
-  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
+# Pivot the data to a wide format
+wide_data1 <- data1 %>%
+  pivot_wider(names_from = candidate_name, values_from = avg_pct) %>%
   rename(Trump = `Donald Trump`, Biden = `Joe Biden`)
 
-merged_data <- map_data |>
-  left_join(wide_data, by = "state")
+# Merge map and polling data
+merged_data1 <- map_data %>%
+  left_join(wide_data1, by = "state")
 
-merged_data <- merged_data |>
+# Categorize color based on polling results
+merged_data1 <- merged_data1 %>%
   mutate(color_category = case_when(
@@
 
-ggplot(data = merged_data) +
+ggplot(data = merged_data1) +
   geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) +
@@
     title = "Voter Sentiment During Biden's Presidential Campaign",
-    x = NULL,
-    y = NULL
-  ) +
-  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
-    legend.direction = "horizontal", 
-    legend.title = element_text(hjust = 0.5)
-  )
-```
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-
-swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
-                  'North Carolina', 'Pennsylvania', 'Wisconsin')
-
-map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)
-
-data <- read_csv('../data/raw_data/president_polls.csv')
-
-biden_drop_date <- as.Date("2024-07-21")
-
-data <- data |>
-  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
-  filter(start_date > biden_drop_date) |>
-  filter(pollscore < -1) |>
-  select(state, pct, candidate_name) |>
-  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) |>  
-  filter(state %in% swing_states) |>
-  group_by(state, candidate_name) |>  
-  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")
-
-map_data <- map_data |> 
-  rename(state = NAME)
-
-wide_data <- data |>
-  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
-  rename(Trump = `Donald Trump`, Harris = `Kamala Harris`)
-
-merged_data <- map_data |>
-  left_join(wide_data, by = "state")
-
-merged_data <- merged_data |>
-  mutate(color_category = case_when(
-    Trump > Harris ~ "Donald Trump",
-    Harris > Trump ~ "Kamala Harris",
-    TRUE ~ "Tie" 
-  ))
-
-ggplot(data = merged_data) +
-  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
-  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "Tie" = "gray"), 
-                    name = "Winner") + 
-  labs(
-    title = "Voter Sentiment after Biden Drops out of the Presidential Race",
     x = NULL,
@@
 #| message: false
+#| label: fig-candidates
+#| fig-cap: "Voter Sentiment Before & After Biden's Candidacy Withdrawl"
+data2 <- data %>%
+  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) %>%
+  filter(start_date > biden_drop_date, pollscore < -1) %>%
+  filter(candidate_name %in% c("Donald Trump", "Kamala Harris"), state %in% swing_states) %>%
+  group_by(state, candidate_name) %>%
+  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")
 
+# Pivot the data for post-withdrawal analysis
+wide_data2 <- data2 %>%
+  pivot_wider(names_from = candidate_name, values_from = avg_pct) %>%
+  rename(Trump = `Donald Trump`, Harris = `Kamala Harris`)
+
+# Merge map and polling data for post-withdrawal
+merged_data2 <- map_data %>%
+  left_join(wide_data2, by = "state")
+
+# Categorize color based on polling results post-withdrawal
+merged_data2 <- merged_data2 %>%
+  mutate(color_category = case_when(
+    Trump > Harris ~ "Donald Trump",
+    Harris > Trump ~ "Kamala Harris",
+    TRUE ~ "Tie"
+  ))
+
+# Create the second plot
+ggplot(data = merged_data2) +
+  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) +
+  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "Tie" = "gray"), 
+                    name = "Winner") +
+  labs(
+    title = "Voter Sentiment After Biden Drops Out of the Presidential Race",
+    x = NULL,
+    y = NULL
+  ) +
+  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
+    legend.direction = "horizontal", 
+    legend.title = element_text(hjust = 0.5)
+  )
+```
+
+#### Poll End Date
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-end-date
+#| fig-cap: "Voter Sentiment Across Time For Donald Trump and Kamala Harris" 
 data <- read_csv('../data/raw_data/president_polls.csv')
@@
   labs(
-    title = "Time Series of Average Polling Percentages",
-    x = "Date",
-    y = "Average Percentage",
-    color = "Candidate"
+    title = "",
+    x = "",
+    y = "PCT",
+    color = ""
   ) +
@@
   )
-
 ```
-
-# Data Overview
 
@@
 
-# Data Cleaning
-
 To clean this data, we first filtered out all hypothetical polls.
@@
 
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-Talk way more about it.
+## Data Cleaning
 
@@
 #| message: false
+#| label: model
+#| fig-cap: "Probability Distribution of Donald Trump's Electoral College Seat Proportion"
 library(tidyverse)
@@
     geom_line(size = 1) + 
-    labs(title = "Probability Distribution of Donald Trump's Electoral College Seat Percentage",
+    labs(title = "",
          x = "Proportion",
@@
 
-# Appendix 1 {-}
+# Appendix 1
 
@@
 
-# Appendix 2 {Idealized Survey}
+# Appendix 2 : Idealized Survey
 
@@
 
-# 2.1 Sampling
+## Sampling
 
@@
 
-# 2.2 Survey Method
+## Survey Method
 
@@
 
-# 2.3 Recruitment
+## Recruitment
 
@@
 
-# 2.4 Data Validation
+## Data Validation
 
@@
 
-# 2.5 Poll Aggregation and Forecasting
+## Poll Aggregation and Forecasting
 
@@
 
-# 2.6 Survey Example
+## Survey Example
 
@@
 
-# Additional Data Details
-
-# Model Details {#sec-model-details}
+# Appendix 3 : Model Details
 
@@
 #| message: false
-
+#| label: alternate-model
+#| fig-cap: "Distribution of Donald Trump's Electoral College Seat Proportion"
 params <- data.frame(
@@
   labs(
-    title = "Distribution of Donald Trump's Electoral College Seat Proportion",
+    title = "",
     x = "Probability",

### 2024-11-04 paper/paper.qmd
fixed subtitle

@@
 title: "Forecasting the 2024 US General Election"
-subtitle: "Analyzing the influence state on election outcome by using 538 Pollster Data with Bayesian hierarchical model, Predicting Harris Victory[need more details in result]"
+subtitle: "Analyzing the Impact of Swing States on Election Outcomes Using 538's 2024 General Election Dataset"
 author: 
@@
 general election.
-Using the poll of polls approach, a generalized linear model was developed with data from FiveThirtyEight [more spcific on method, and what is the date cutoff of the data]  The model's prediction is based on the popular vote in the seven key swing states—Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner.
-Results indicate that Harris is predicted to win, by [how much?
-need to be more spcific on the result].
-This suggests that current polling trends may provide reliable insights into the likely election result, though future shifts in voter sentiment could impact the final outcome."
+Using the poll of polls approach, a bayesian hierarchical model was developed with data from FiveThirtyEight using polling data from July 21 to October 19, 2024.
+The model's prediction is based on the electoral college vote in the seven key swing states : Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner.
+Results indicate that Harris is predicted to win with a probaility of 66%."
 format: pdf

### 2024-11-04 scripts/00-simulate_data.R
simulation

@@ -31,5 +31,10 @@ data <- data.frame(
 )
 
 data$pct <- pmin(pmax(data$pct, 0), 100)
+data$pollscore <- runif(1000, -1.5, -0.4)
+data$candidata_name <- sample(c("Trump", "Harris"), 1000, replace = TRUE)
+data$sample_size <- sample(500:10000, 1000, replace = TRUE)
+data$end_date <- as.Date("2020-01-01") +
+ sample(0:(as.Date("2024-10-31") - as.Date("2020-01-01")), 1000, replace = TRUE)
 
 write_csv(data, './data/simulated_data/simulated_data.csv')

### 2024-11-04 paper/paper.qmd
refine model section
Update paper.qmd
minor change in titel and abstract

@@
 title: "Forecasting the 2024 US General Election"
-subtitle: "Analyzing the Impact of Swing States on Election Outcomes Using 538's 2024 General Election Dataset"
+subtitle: "Analyzing Swing State Impact on 2024 Election Outcomes: Bayesian Hierarchical Model Using 538's Data Predicts 66% Harris Victory"
 author: 
@@
 general election.
-Using the poll of polls approach, a bayesian hierarchical model was developed with data from FiveThirtyEight using polling data from July 21 to October 19, 2024.
-The model's prediction is based on the electoral college vote in the seven key swing states : Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner.
-Results indicate that Harris is predicted to win with a probaility of 66%."
+Using the poll of polls approach, a Bayesian hierarchical model was developed with data from FiveThirtyEight using polling data from July 21 to October 19, 2024.
+The model's prediction is based on the electoral college vote in the seven key swing states: Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner.
+Results indicate that Harris is predicted to win with a probability of 66%."
 format: pdf
@@
 Biden election.
-A Bayesian updating and Monte Carlo Simulations is created to predict the winner of the election outcome based on the popular vote across these swing states[some more details for what was done?].
-The primary estimand in this analysis is the probability distribution of Harris's win percentage, derived from current polling data and historical election results.
-The model predicted that Harris will win, and [add details to result]
+A Bayesian updating and Monte Carlo Simulations are created to predict the winner of the election outcome based on the popular vote across these swing states[some more details for what was done?].
+The estimand in this analysis is the probability distribution of Harris's win percentage, derived from current polling data and historical election results.
+The model predicted that Harris would win, and [add details to result]
 
@@
 
-## Posterior Predictive Check
-
 # References

### 2024-11-04 paper/paper.qmd
updated introduction

@@
 
-The 2024 United States presidential election, the 60th quadrennial election, is scheduled for November 5, 2024.
+The 2024 United States presidential election, is scheduled for November 5, 2024.
 The stakes are high, as the outcome will shape domestic policies and impact international relations, with effects on trade, climate change, and security.
@@
 
-The primary candidates in this election, Democratic incumbent Kamala Harris and former Republican president Donald Trump, are expected to engage in a tight race, with polling indicating that the results could be very close.
+The primary candidates in this election, Democratic vice president Kamala Harris and former Republican president Donald Trump, are expected to engage in a tight race, with polling indicating that the results could be very close.
 Swing states—those that do not consistently vote for one party—are vital in determining the overall outcome, as their electoral votes can sway the final result.
-Seven important swing states—Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina—are expected to be battlegrounds in this election, making their polling data essential for accurate forecasts.
+This year's swing states are: Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina.
+They expected to be battlegrounds in this election, making their polling data essential for accurate forecasts.
 
-This paper focuses on forecasting the election outcome by employing the "poll of polls" method, utilizing data from FiveThirtyEight, along with historical data from the 2016 Clinton vs.
-Trump election and the 2020 Trump vs.
-Biden election.
-A Bayesian updating and Monte Carlo Simulations are created to predict the winner of the election outcome based on the popular vote across these swing states[some more details for what was done?].
-The estimand in this analysis is the probability distribution of Harris's win percentage, derived from current polling data and historical election results.
-The model predicted that Harris would win, and [add details to result]
+This paper focuses on forecasting election outcomes by employing the "poll of polls" method, utilizing data from FiveThirtyEight.
+A Bayesian hierarchical model and Monte Carlo Simulations have been used to predict the winner of the election based on the electoral college vote, and quantify how close the election actually is.
+The estimand in this analysis is the proportion of seats Donald Trump will win in the 2024 election.
+Using the estimand we can conclude election results.
+Assuming Donald Trump's loss is Kamala Harris' win, she has around a 66% chance of winning this election.
 
@@
 Ultimately, the findings from this study can improve the accuracy of predictions in a political landscape characterized by rapid change and polarization, thereby enhancing the relevance and reliability of electoral analyses.
-
-The remainder of this paper is structured as follows: .......
 

### 2024-11-04 paper/paper.qmd
sample sizes

@@
 The most common sample size of polls is around 1250, as seen in @fig-sample-size.
+Although this may seem like a small amount, sample sizes greater than 1004 people have a margin of error less than 3 percent @citeScientificAmerican.
+Standard political polls usually aim to include more than 1000 people.
+This is a good indicator that most of the polls in our data follow the accepted pratices of polling.
+
+The sample size of a poll refers to the number of individuals who participate in the survey.
+Generally, a larger sample size is considered to provide a more accurate representation of the overall population.
+For instance, a common sample size for polls is greater than 1000.
+While this might seem like a relatively small number, studies have shown that sample sizes exceeding 1,004 can achieve a margin of error of less than 3 percent (Scientific American).
+Sample sizes in our dataset have a mode of 1250, as seen in @fig-sample-size, which suggests that most of the polls in the dataset adhere to accepted polling practices when it comes to sample size.
+This commitment to robust sample sizes enhances the reliability of the polling data we analyze.
 

### 2024-11-04 paper/paper.qmd
more refinement on the data section

--- 2024-11-04 paper/paper.qmd
@@
 ### Sample Size
-The sample size of a poll is the number of people who participated in the poll.
-Generally, a larger sample size is seen as a better representation of the population.
-The most common sample size of polls is around 1250, as seen in @fig-sample-size.
+The sample size of a poll refers to the number of individuals who participated in the poll.
+Generally, a larger sample size is considered to provide a more accurate representation of the overall population.
+For instance, a common sample size for polls is greater than 1000.
+While this might seem like a relatively small number, studies have shown that sample sizes exceeding 1,004 can achieve a margin of error less than 3 percent,  @citeScientificAmerican.
+Sample sizes in our dataset have a mode of 1250, as seen in @fig-sample-size, which suggests that most of the polls in the dataset adhere to accepted polling practices when it comes to sample size.
 

### 2024-11-04 paper/paper.qmd
candidates write up

@@
 The pollscore feature serves as a metric for evaluating the quality of a poll.
-It is a mesaurement that tries to quantify the realiablity of a poll.
+It is a mesaurement that tries to quantify the reliablity of a poll.
 It is calculated based on factors such as poll transparency, bias, and error, @citeABC.
@@
 ### Percentage of Votes
-Each poll aims to .
-By examining how these values fluctuate over time and in response to different candidates, we can better understand their impact on public opinion.
-Currently, national level polls show Kamala Harris with a 46% approval rating compared to Donald Trump's 43%.
-This highlights the importance of candidate dynamics in shaping voter sentiment across the electorate.
+Each observation in the dataset (poll) has a PCT.
+This is the percentage of votes that a particular candidate recieved in the poll.
+The PCT attempts to measure voter sentiment in the United States and predict election outcomes using this quantified voter sentiment.
+PCT is distributed differently based on the candidate and the state of the poll.
+PCT values can also shift over time indicating a shift in voter sentiment.
+Currently, national level polls show that Kamala Harris has an approval rating of 48% while Donald Trump's approval rating is 46%, @cite538.
+Looking at trends in vote percentages will be crucial for data cleaning, and preventing model overfitting.
+
+\newpage
 
 #### Candidates
+The United States has a strong two-party system, consisting of the Liberal Democratic Party and the Conservative Republican Party.
+At the beginning of the year, the primary candidates for the presidential race were current president Joe Biden representing the Democratic Party, and former president Donald Trump representing the Republican Party.
+However, after Biden withdrew from the race on July 21st, Kamala Harris emerged as the new front-runner for the Democratic Party.
+
+To analyze general voter sentiment in swing states for each candidate, we can examine the percentage of votes (PCT) associated with each party's main candidate in those states before and after Biden's withdrawal.
+A Democratic majority is represented in blue, while a Republican majority is depicted in red.
+As seen in @fig-inital-candidates, during Joe Biden's presidential run, Donald Trump had a higher average PCT.
+Notably, following Biden's exit, the Democratic Party seems to have gained increased support in key swing states such as Wisconsin, Michigan, and Pennsylvania, indicating a significant shift in voter sentiment toward Kamala Harris, @fig-candidates-after.
+
 ```{r}
@@
 #| message: false
+#| label: fig-inital-candidates
+#| fig-cap: "Voter Sentiment in Swing States During Joe Biden's 2024 Presidential Campaign"
 library(sf)
@@
   labs(
-    title = "Voter Sentiment During Biden's Presidential Campaign",
+    title = "",
     x = NULL,
@@
 #| message: false
-#| label: fig-candidates
-#| fig-cap: "Voter Sentiment Before & After Biden's Candidacy Withdrawl"
+#| label: fig-candidates-after
+#| fig-cap: "Voter Sentiment in Swing States After Biden's Candidacy Withdrawl"
 data2 <- data %>%
@@
   labs(
-    title = "Voter Sentiment After Biden Drops Out of the Presidential Race",
+    title = "",
     x = NULL,
@@
 
+\pagebreak
 #### Poll End Date

### 2024-11-04 paper/paper.qmd
update pdf

@@
 
+The timing of a poll is very important to it's accuracy as voter sentiment can change a lot with things such as scandals, new economic conditions, natural disasters, etc.
+To build our model we need to use relevant recent data as well as make sure that there's is enough data to analyze.
+@fig-end-date shows trends in national voter PCT across time after Joe Biden dropped out of the race.
+
 ```{r}
@@
 ```
-
-Our data is the Project 538 [cite] Presidential General Election polls dataset (current cycle) from October 17th.
-The data contains roughly 16000 observations of polls taken from January 2023 at the earliest to October 16th at the latest.
-Each poll has been given an associated pollscore, numeric grade and transparency score, and the methodology used to conduct it is also given.
-We also know the duration of the poll by start and end date, as well as the time of publication and the number of particpants.
-The candidate sponsoring each poll is given as well as the candidate endorsed by each poll.
-
-(can be deleted)
-Each poll has an associated pollster, and each pollster is assigned a numeric grade (0.5 to 3.0) and pollscore (-1 to 1) corresponding to the quality and reliability of the pollster and a transparency score (0 to 10) corresponding to the level of transparency the pollster provides in regards to their methodology.
-Furthermore, each poll has an associated methodology.
-[graphics to include?, pollscore vs numeric_grade and numeric_grade vs transparency score can show strong linear relationship, providing justification to use one]
-
-Important to our modeling are the variables answer and pct, where answer indicates what candidate the poll predicts will receive the associated percentage (pct) of votes.
-Thus each poll is captured at least twice in the dataset, once for each polled candidate.
-The associated unique poll_id is a variable that then captures multiple observations of the same poll.
-Another important variable is Hypothetical, which notes whether a poll is between candidates that ran against each other in reality, or in a hypothetical scenario.
-Another important variable for our modeling is state, which indicates which state, if any, the poll was focused.
-[pct by state for both candidates can show that swing states are less clear]
-
-To clean this data, we first filtered out all hypothetical polls.
-Then, we filtered out any polls not focused on swing states.
-Next,  we looked only at polls answering a percentage of Trump as he appeared in the last two presidential races, and we removed all data from from before Biden dropped out to get results for Trump vs Harris.
-
-Next, we wanted to focus only on high quality polls, so we filtered out all polls with a pollscore greater than zero.
 

### 2024-11-04 paper/paper.qmd
Update cross reference

@@
 
-The primary candidates in this election, Democratic vice president Kamala Harris and former Republican president Donald Trump, are expected to engage in a tight race, with polling indicating that the results could be very close.
+The primary candidates in this election, Democratic Vice President Kamala Harris and former Republican president Donald Trump, are expected to engage in a tight race, with polling indicating that the results could be very close.
 Swing states—those that do not consistently vote for one party—are vital in determining the overall outcome, as their electoral votes can sway the final result.
@@
 Ultimately, the findings from this study can improve the accuracy of predictions in a political landscape characterized by rapid change and polarization, thereby enhancing the relevance and reliability of electoral analyses.
+
+The remainder of this paper is structured as follows: @sec-data-overview discusses the data sources and variables used in the analysis; #sec-Dataset-Features presents the visualization of results; @sec-data-cleaning explains the process of cleaning the data; @sec-model justifies the model setup, as well as assumptions and limitations; @sec-result presents the result of the models; @sec-discussion provides a discussion of the paper; @sec-appendix-one analyzed the methodology of The New York Times/Siena College pollster; @sec-appendix-two created an ideal survey of the poll.
+@sec-appendix-three provides a more detailed explanation of the models.
+Finally, @sec-reference provides a list of references.
 
@@
 
-## Data Cleaning
+## Data Cleaning {#sec-data-cleaning}
 
-# Model
+# Model {#sec-model}
 ## Model Justification
@@
 
-# Results
+# Results {#sec-result}
 
@@
 
-# Discussion
+# Discussion {sec-discussion}
 
@@
 
-# Appendix 1
+# Appendix 1 {#sec-appendix-one}
 
@@
 
-# Appendix 2 : Idealized Survey
+# Appendix 2 : Idealized Survey {#sec-appendix-two}
 
@@
 
-# Appendix 3 : Model Details
+# Appendix 3 : Model Details {#sec-appendix-three}
 
@@
 
-# References
+# References {#sec-reference}

### 2024-11-04 paper/paper.qmd
Update paper.qmd

@@
 
-## First discussion point {#sec-first-point}
+## Overview and What Our Results Can Imply
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+In this paper, we took a dataset of US presidential election polls and considered a sample of high-quality polls and the prediction they gave for Trump in the Trump vs Harris race.
+We focused our poll of polls model on this election’s swing states as we have a general overview of the number of seats both Harris and Trump will win outside of them and created a Bayesian model to estimate the probability of Trump being elected by estimating the probability distribution of the proportion of electoral college seats he will hold.
+We started with Beta priors chosen by analyzing Trump’s performance in the prior two elections.
+Then, we simulated 10000 elections using Monte Carlo Simulations, using our polling data to determine the Swing States Trump would win in each simulated election and used this data to create a posterior distribution for Trump’s proportion of seats.
+The mean of this distribution is our estimated average election outcome for the percentage of seats Trump will capture.
+The mean and shape of the updated beta distribution are similar to a normal distribution.
+In fact, it resembles the distribution of the probability of a coin flip for a large number of observations with its mean of ~0.45 compared to the coin flip’s 0.50, and with similar tails.
+This points to high uncertainty.
+In this discussion, we want to raise a question, how often can we expect to see a distribution like this, and what, if anything can we say about the population’s voting come election day?
 
-## Second discussion point
+The United States has long had an outright monopoly over electoral and congressional seats between the Democratic and Republican parties, what this distribution shows is that among registered voters who answered polls (a small, and somewhat atypical group), the margins are close within Swing States, with a slight Harris lead.
+This, however, doesn’t feel like a novel discovery, we already know that swing states do not have a consistent voting pattern.
+What we have done in this paper then is reinforce the notion that this election will be close, which adds us to the chorus of pollsters singing a similar song.
+As noted in our model section, we assumed that our Trump priors would affect our result, a reasonable assumption as people’s voting patterns do not tend to change drastically.
+Moreover, we must assume that the Monte Carlo simulations we made with the chosen learning rate are representative, which relies on the assumption that the polling data we used is reflective of reality, which is a necessary assumption and the most vital one when we look to extrapolate results about reality from a sample, which is the goal of surveying and statistics more generally.
+There is always reason for concern and uncertainty, but we can do nothing but trust in and maintain good statistical practice.
 
-## Third discussion point
+In conclusion, we can say only that per our data, we believe that the polls reflect that if the election were held on October 17th 2024, we expect to see Trump holding 44% of the electoral seats
 
-## Weaknesses and next steps
+## Polarization in the US and accuracy of polling results
 
-Weaknesses and next steps should also be included.
+We estimated 44% as Trump’s election rate, with a similar distribution to that of a coin flip, indicating high uncertainty.
+This uncertainty is not unwarranted, most other polls and aggregates are predicting this to be a close election for a number of reasons.
+We believe that the primary one is, that the social, cultural, and by extent, political landscape, of the United States has changed in the past four years.
+Saying that people are divided is nothing new, it has been a clear observation for some years now and the media have been documenting it since before Trump’s election in 2016.
+Economic uncertainty and general public unease make for a political landscape that is not as easily definable as it was in some previous eras of the United State’s existence.
+Economic inequality has risen to Great Depression levels [], with people being financially burdened to an untenable degree; people’s immediate futures are uncertain with homelessness becoming a more and more apparent threat for many.
+This breeds radicalization in various avenues as people look for answers and stability, with reactionary opportunists eager to satisfy.
+As this reactionary sentiment has grown into a new alt-right, it has brought legitimate political power to its leaders.
+Anti-immigration policies have only grown in popularity, Trump’s “Build The Wall” slogan is now replaced with a notion that Haitian immigrants and descendants are harming domestic animals and turning to crime in the Midwest.
+LGBT rights are being suppressed in parts of the country, with DeSantis’ anti-trans bill in Florida being a prime example.
+On the other side of the aisle, the staunchly status-quo democrats have done little to inspire hope for change, and in the face of said growing uncertainty it leaves people feeling shaken in their beliefs, and as if they are without the support of their institutions.
+Of course, why should they feel supported?
+Hurricane Helene just devastated the eastern United States and many people were left almost entirely without support for days and weeks under a Democrat administration, Roe v Wade was overturned in the same administration, does this inspire belief and hope?
+Without even mentioning the radicalization happening on both sides in the wake of BLM, COVID-19, the first easily visible effects of climate change for Americans, the Ukraine War, and the new Israeli cleansings and conflicts in Palestine and Lebanon, it is clear to see that the United States is in a new social paradigm, and people are looking for solutions.
 
-\newpage
+When in a new paradigm, can we expect accurate polling?
+The notion of Shy Tories may be weak, but what does throw a wrench in polling results is the higher-than-expected voter turnout for this race.
+[] We just spent a section discussing how divided and polarized people have been in the country.
+Pew Research Center notes that some 76% of voters believe the election outcome is crucial in 2024 and with non-response bias among Republicans having been a concern in prior elections, there is reason to believe that the polls may be inaccurate.
+However, polling is not done sparsely, and many polls track changes in sentiment, Pew Research Centre is showing changes in key issues as being in line with the general public discussion, so perhaps the pollsters aren’t being blindsided [].
+Moreover, higher political mobilization and higher polarization might mean that there is more motivation for people to voice their beliefs in polls for an election they believe is as important as it is.
 
-\appendix
+We must also consider the election strategies of both candidates and whether they are maximizing the votes their base of support can bring in, and whether they are growing their bases as well.
+We know many arab voters who are historically Democrats, have turned to the Republican side as many have lost hope in the Democrats to end the war in Gaza.
+This is a simple point to the increased polarization of the country, and Democrats are not shifting policy-wise to meet voters.
+Harris already faces challenges as a first-time candidate, and a woman of colour, the Democratic party is not helping by having a much less easily definable policy than the Republicans.
+Despite this, the Dems remain more popular in polls.
+Of course, non-response bias is always a concern, but demographics like Michigan Arabs have given a clear polling perspective of moving right.
+
+So a key question becomes “Is it a close race due to poor polling assumptions producing uncertain polling results, or due to it truly being a close race?” We can’t know for certain until we know, but if we are to trust the polling data, it will be a close race.
+An important point to make is one about echo chambers.
+Pollsters are often criticized for publishing outlier results, which can produce bias as many pollsters converge to a consensus.
+As stated in the previous discussion, our individual findings are part of this general consensus, which could be a point of concern if we weren’t confident in our priors and our simulation.
+We reach a similar conclusion as before, we can only trust in our data and techniques.
+
+## Weaknesses and Next Steps
+
+As has been stated in our model assumptions, this paper does not account for any shifts that occurred after October 16th, 2024, which is critical as the weeks and days leading up to an election are the most important in determining the outcome.
+The model also says nothing of Harris’s seats, only Trump’s.
+We do not expect third parties to hold a number of seats above 3-4%, which would still signal a Harris victory, however, her results were not modelled as Trump’s were which does provide a blindspot.
+
+As for next steps, if we were to repeat our process for other election cycles, would we expect to see a similar distribution when swing states are as close as they have been for this race?
+Taking this direction could allow us to speak about the impact of Swing States more broadly.
+
+Additionally, producing a model for Harris and comparing it to Trump would cover a key weakness of our paper and allow for a more complete picture of voting sentiment than we have now.
 
